---
title: |
    | URD: try on full dataset
    | Quick start like
author: "Céline Chevalier"
date: format(Sys.time(), "%D")
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
params:
    sub:
        value: 1
    knn:
        value: 200
    sigma:
        value: "NULL"
    floodNsim:
        value: 10
    cellsForward:
        value: 20
    cellsBack:
        value: 40
    RWmaxSteps:
        value: 50
    treeMeth:
        value: preference
        choices: [preference, ks]
    cellsPerPTbin:
        value: 20
    binsPerPTwindow:
        value: 5
    treePtreshold:
        value: 0.001
    seed:
        value: 17
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    })
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    })
  })
</script>


```{r setup, warnings=F}
# Set library to rely only on the urd environment
.libPaths(c(.libPaths()[2]))

# Load packages
suppressPackageStartupMessages(library(rgl))
suppressPackageStartupMessages(library(URD))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kableExtra))

attach(params)
PATH_FIGURES <- paste0("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/fullDataset_lastYearImpl/figures/eureka4/")
if(!dir.exists(PATH_FIGURES)){ dir.create(PATH_FIGURES) }

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.path = PATH_FIGURES,
    dev = c('png','pdf')
    )
rgl::setupKnitr()
```


```{r show-params}
ul <- unlist(params)
df <- data.frame(keyName=names(ul), value=ul, row.names=NULL)
names(df) <- c('Parameters', 'Values')
df1 <- df[1:ceiling(dim(df)[1]/2),]
df1$Parameters <- cell_spec(df1$Parameters, bold = T)
df2 <- df[(ceiling(dim(df)[1]/2)+1):dim(df)[1],]    #(2*ceiling(dim(df)[1]/3)),]
df2$Parameters <- cell_spec(df2$Parameters, bold = T)
# df3 <- df[((2*ceiling(dim(df)[1]/3))+1):dim(df)[1],]
# df3$Parameters <- cell_spec(df3$Parameters, bold = T)


# list(df1, df2, df3) %>%
list(df1, df2) %>%
  knitr::kable(align = "ll", row.names = FALSE, escape = F) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
rm(df)
```

# Load the data

For this analysis, I use the merged data of the 4 days (ie. lab_4_days_merged). I get an urd-object by running the
following code in bash, under the rossiV4 conda environment.

```{bash, eval=FALSE}
Rscript ./exportToURD.R
```

```{r loadData}
# Load the 4 days merged dataset
object <- readRDS(file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/inputData/lab_data/urdObject_lab_4_days.rds")
```


# Calculate variable features

CV: coefficient of variation

```{r CalcVariableFeatures}
stages <- sort(unique(object@group.ids$day))
cells.each.stage <- lapply(stages,
                           function(stage){
                               rownames(object@group.ids)[which(object@group.ids$day == stage)]
                           }
)

var.genes.by.stage <- lapply(1:length(stages),
                             function(n){
                                 findVariableGenes(object,
                                                   cells.fit = cells.each.stage[[n]],
                                                   set.object.var.genes = F,
                                                   diffCV.cutoff = 0.001,
                                                   mean.min = 0.001,
                                                   mean.max = 100,
                                                   main.use = stages[[n]],
                                                   do.plot = T
                                 )
                             }
)

# combine variable features
names(var.genes.by.stage) <- stages
var.genes <- sort(unique(unlist(var.genes.by.stage)))
length(var.genes)
object@var.genes <- var.genes
```

# Calculate PCA and tSNE

```{r CalcPCA}
set.seed(seed)
object <- calcPCA(object, mp.factor = 2)
pcSDPlot(object)
```

```{r CalcTsne}
# Remove cells with duplicates PCA coordinates
cellsToRemove <- rownames(object@pca.scores[duplicated(object@pca.scores),])
print(cellsToRemove)
object <- urdSubset(object,
                    cells.keep = setdiff(colnames(object@logupx.data), cellsToRemove)
)

# Perform Tsne projection
set.seed(seed)
object <- calcTsne(object,
                   perplexity = 30,
                   theta=0.5
)

plotDim(object, "day", plot.title = "tSNE: Day")
```


# Remove outliers

```{r rmOutliers}
# Remove outliers
# Calculate a k-nearest neighbor graph
set.seed(seed)
object <- calcKNN(object, nn=knn)
outliers <- knnOutliers(object,
                        nn.1=1,
                        nn.2=20,
                        x.max=10,
                        slope.r=1.4,
                        int.r=1.5,
                        slope.b=0.8,
                        int.b=4.6,
                        title = paste0("Identifying Outliers by k-NN Distance")
)

# Subset object to eliminate outliers
print(paste0("Number of removed outliers: ", length(outliers)))
cells.keep <- setdiff(colnames(object@logupx.data), outliers)
object <- urdSubset(object, cells.keep=cells.keep)
```

# Calculate diffusion map (DM)

```{r CalcDM}
root.cells <- rownames(object@meta)[object@meta$day=="Day_04"]
if (sigma == "NULL"){
    set.seed(seed)
    object <- calcDM(object,
                     knn=knn,
                     dcs.store = 50
    )
} else {
    set.seed(seed)
    object <- calcDM(object,
                     knn=knn,
                     dcs.store = 50,
                     sigma.use=sigma
    )
}

colors.use <- rainbow(length(unique(object@group.ids$day)))
plotDimArray(object = object,
             reduction.use = "dm",
             dims.to.plot = 1:18,
             label="day",
             plot.title="",
             outer.title=paste0("Day - Diffusion Map Sigma ", sigma, "\n"),
             legend=F,
             alpha=0.3,
             discrete.colors=colors.use
)
```

# Calculate pseudotime (PT)

```{r CalcPT}
set.seed(seed)
flood.result <- floodPseudotime(object,
                                root.cells=root.cells,
                                n=floodNsim,
                                minimum.cells.flooded=2,
                                verbose=T
)

set.seed(seed)
object <- floodPseudotimeProcess(object,
                                 flood.result,
                                 floods.name="pseudotime",
                                 max.frac.NA=0.4,
                                 pseudotime.fun=mean,
                                 stability.div=20
)

pseudotimePlotStabilityOverall(object)
plotDim(object, "pseudotime", plot.title = "Pseudotime")
plotDists(object, "pseudotime", "day")
```

# Define tips

```{r tipsSubset}
tips_cells <- cellsInCluster(object, "day", "Day_11")
renamed_tips <- gsub("11_", "", object@group.ids[rownames(object@group.ids) %in% tips_cells, 'singleTP_res1'])
object@group.ids[rownames(object@group.ids) %in% tips_cells, "tip.clusters"] <- renamed_tips

# Cluster cells from final stage
plotDim(object, "tip.clusters", plot.title = "Leiden clusters -\nfrom Seurat analysis", legend = FALSE) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())
```

# Biased random walks

```{r performPTlogistic}
set.seed(seed)
object.ptlogistic <- pseudotimeDetermineLogistic(object,
                                                 "pseudotime",
                                                 optimal.cells.forward=cellsForward,
                                                 max.cells.back=cellsBack,
                                                 do.plot = T
)

# Bias the transition matrix according to pseudotime
set.seed(seed)
object.biased.tm <- as.matrix(pseudotimeWeightTransitionMatrix(object,
                                                               "pseudotime",
                                                               logistic.params=object.ptlogistic
))

# Simulate the biased random walks from each tip
set.seed(seed)
object.walks <- simulateRandomWalksFromTips(object,
                                            tip.group.id="tip.clusters",
                                            root.cells=root.cells,
                                            transition.matrix = object.biased.tm,
                                            max.steps = RWmaxSteps,
                                            n.per.tip = 2 * length(tips_cells),
                                            verbose = F
)

# Process the biased random walks into visitation frequencies
set.seed(seed)
object <- processRandomWalksFromTips(object, object.walks, verbose = F)

plotDim(object, "tip.clusters", plot.title="Cells in each tip")

# saveRDS(object, file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/fullDataset_lastYearImpl/rdsObject/eureka4_object.rds")
```

# Build tree

```{r buildTree}
# Load the cells used for each tip into the URD object
object.tree <- loadTipCells(object, "tip.clusters")
# Build the tree
set.seed(seed)
object.tree <- buildTree(object.tree,
                         tips.use = levels(as.factor(object.tree@group.ids$tip.clusters)),
                         pseudotime = "pseudotime",
                         divergence.method = treeMeth,
                         cells.per.pseudotime.bin = cellsPerPTbin,
                         bins.per.pseudotime.window = binsPerPTwindow,
                         save.all.breakpoint.info = T,
                         p.thresh = treePtreshold,
                         min.cells.per.segment = 0
)

plotTree(object.tree, title="Labelled segments", tree.size = 0.3, label.segments = T) +
    coord_flip()

plotTree(object.tree, "day", title="Developmental Stage", tree.size = 0.3) +
    coord_flip() +
    theme(plot.title = element_text(hjust = 0.5))
```

<!-- # Save object -->

<!-- ```{r, eval=FALSE} -->
<!-- saveRDS(object.tree, file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/fullDataset_lastYearImpl/rdsObject/eureka4_tree.rds") -->
<!-- ``` -->

