---
title: "URD: trajectory inference"
author: "Céline Chevalier"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
params:
    knn:
        value: 60
    sigma:
        value: 3
    floodNsim:
        value: 5
    cellsForward:
        value: 40
    cellsBack:
        value: 80
    RWmaxSteps:
        value: 50
    RWNsim:
        value: 10000
    treeMeth:
        value: preference
        choices: [preference, ks]
    cellsPerPTbin:
        value: 25
    binsPerPTwindow:
        value: 8
    treePtreshold:
        value: 0.001
    seed:
        value: 17
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->

<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->


<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    })
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    })
  })
</script>

```{r setup, warnings=F}
# Set library to rely only on the urd environment
.libPaths(c(.libPaths()[2:1]))

# Load packages
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(URD))
suppressPackageStartupMessages(library(rgl))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(DT))

source("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/gastruloid_timeserie_scRNA-seq/scripts/utilities/additionalPlots.R")

attach(params)
PATH_FIGURES <- paste0("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/paramOptim/figures/",
                       "knn.", knn,
                       "_sigma.", sigma,
                       "_floodNsim.", floodNsim,
                       "_cellsForward.", cellsForward,
                       "_cellsBack.", cellsBack,
                       "_RWmaxSteps.", RWmaxSteps,
                       "_RWNsim.", RWNsim,
                       "_treeMeth.", treeMeth,
                       "_cellsPerPTbin.", cellsPerPTbin,
                       "_binsPerPTwindow.", binsPerPTwindow,
                       "_treePtreshold.", treePtreshold, "/"
)

PATH_RDS <- paste0("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/paramOptim/rdsObjects/",
                       "knn.", knn,
                       "_sigma.", sigma,
                       "_floodNsim.", floodNsim,
                       "_cellsForward.", cellsForward,
                       "_cellsBack.", cellsBack,
                       "_RWmaxSteps.", RWmaxSteps,
                       "_RWNsim.", RWNsim,
                       "_treeMeth.", treeMeth,
                       "_cellsPerPTbin.", cellsPerPTbin,
                       "_binsPerPTwindow.", binsPerPTwindow,
                       "_treePtreshold.", treePtreshold, "/"
)

if(!dir.exists(PATH_FIGURES)){ dir.create(PATH_FIGURES) }
if(!dir.exists(PATH_RDS)){ dir.create(PATH_RDS) }

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.path = PATH_FIGURES,
    dev = c('png','pdf')
    )
rgl::setupKnitr()
```

```{r show-params}
ul <- unlist(params)
df <- data.frame(keyName=names(ul), value=ul, row.names=NULL)
names(df) <- c('Parameters', 'Values')
df1 <- df[1:ceiling(dim(df)[1]/3),]
df1$Parameters <- cell_spec(df1$Parameters, bold = T)
df2 <- df[(ceiling(dim(df)[1]/3)+1):(2*ceiling(dim(df)[1]/3)),]
df2$Parameters <- cell_spec(df2$Parameters, bold = T)
df3 <- df[((2*ceiling(dim(df)[1]/3))+1):dim(df)[1],]
df3$Parameters <- cell_spec(df3$Parameters, bold = T)

list(df1, df2, df3) %>%
    knitr::kable(align = "ll", row.names = FALSE, escape = F) %>% 
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE)
rm(df)
```

```{r setup2, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  echo = TRUE
)
```

# Load the data

For this analysis, I use the merged data of the 4 days (ie. lab_4_days_merged). I get an urd-object by running the following code in bash, under the rossiV4 conda environment.

```{bash, eval=FALSE}
Rscript ./exportToURD.R
```

```{r loadData}
# Load the 4 days merged dataset
object <- readRDS(file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/inputData/lab_data/urdObject_lab_4_days.rds")
```

# Calculate variable features

CV: coefficient of variation

```{r CalcVariableFeatures}
stages <- sort(unique(object@group.ids$day))
cells.each.stage <- lapply(stages,
                           function(stage){
                               rownames(object@group.ids)[which(object@group.ids$day == stage)]
                           }
)

var.genes.by.stage <- lapply(1:length(stages),
                             function(n){
                                 findVariableGenes(object,
                                                   cells.fit = cells.each.stage[[n]],
                                                   set.object.var.genes = F,
                                                   diffCV.cutoff = 0.001,
                                                   mean.min = 0.001,
                                                   mean.max = 100,
                                                   main.use = stages[[n]],
                                                   do.plot = T
                                 )
                             }
)

# combine variable features
names(var.genes.by.stage) <- stages
var.genes <- sort(unique(unlist(var.genes.by.stage)))
print(paste0("Number of variable genes: ", length(var.genes)))
object@var.genes <- var.genes
```

# Load UMAP coordinates

```{r loadUMAP}
umapCoord <- read.table("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/ciml5/lab_4_days_merged/final_umap2Dcoordinates_lab_4_days_merged.csv",
                        sep = ",",
                        header = TRUE)
object@tsne.y <- umapCoord
names(object@tsne.y) <- c("tSNE1", "tSNE2")

plotDays <- plotDim(object, reduction.use = "tsne", 'day', plot.title = "") +
    xlab('UMAP_1') +
    ylab('UMAP_2')
plotDays + theme(legend.position = "none")

grid.newpage()
grid.draw(g_legend(plotDays))
```

# Remove outliers

```{r rmOutliers}
# Remove outliers
# Calculate a k-nearest neighbor graph
set.seed(seed)
object <- calcKNN(object, nn=knn)
outliers <- knnOutliers(object,
                        nn.1=1,
                        nn.2=knn,
                        x.max=10,
                        slope.r=1.3,
                        int.r=2.1,
                        slope.b=0.75,
                        int.b=6.2,
                        title = paste0("Identifying Outliers by k-NN Distance")
)

# Subset object to eliminate outliers
print(paste0("Number of removed outliers: ", length(outliers)))

table(object@group.ids[outliers,c("singleTP_celltype", "singleTP_res1")] %>%
          arrange(singleTP_res1) %>%
          arrange(singleTP_celltype)) %>%
    kbl(caption = "Repartition of the outliers among clusters\n(from single time point analysis)") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

cells.keep <- setdiff(colnames(object@logupx.data), outliers)
object <- urdSubset(object, cells.keep = cells.keep)
```

```{r save1}
saveRDS(object, file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/paramOptim/rdsObjects/trimmedObject.rds")
```

```{r setup3, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  # code evaluation
  eval = FALSE,
  echo = FALSE
)
```

```{r load1}
object <- readRDS(file="/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/trajectory/ciml5/paramOptim/rdsObjects/trimmedObject.rds")
```

```{r setup4, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  echo = TRUE
)
```

# Calculate diffusion map (DM)

```{r CalcDM}
root.cells <- rownames(object@meta)[object@meta$day=="Day_04"]
if (sigma == "NULL"){
    if (knn == "NULL"){
        set.seed(seed)
        object <- calcDM(object,
                         dcs.store = 50
        )
    } else {
        set.seed(seed)
        object <- calcDM(object,
                         knn=knn,
                         dcs.store = 50
        )
    }
} else {
    if (knn == "NULL"){
        set.seed(seed)
        object <- calcDM(object,
                         dcs.store = 50,
                         sigma.use=sigma
        )
    } else {
        set.seed(seed)
        object <- calcDM(object,
                         knn=knn,
                         dcs.store = 50,
                         sigma.use=sigma
        )
    }
}
```

```{r saveDM}
saveRDS(object@dm,
        file = paste0(PATH_RDS, "dm_SIGMA.", sigma, "_KNN.", knn, ".rds")
)
```

```{r plotDM_colDays}
colors.day <- rainbow(length(unique(object@group.ids$day)))
plotDimArray(object = object,
             reduction.use = "dm",
             dims.to.plot = 1:18,
             label="day",
             plot.title="",
             outer.title=paste0("Day - Diffusion Map Sigma ", sigma, "\n"),
             legend=F,
             alpha=0.3,
             discrete.colors=colors.day
)

grid.newpage()
grid.draw(g_legend(plotDays))
```

```{r, eval=FALSE, echo=FALSE}
plotDim3D(
  object,
  'day',
  label.type = "search",
  reduction.use = c("dm"),
  view = NULL,
  dim.1 = 1,
  dim.2 = 2,
  dim.3 = 9,
  w.1 = NULL,
  w.2 = NULL,
  w.3 = NULL,
  cells = NULL,
  alpha = 0.05,
  size = 4,
  title = NULL,
  title.cex = 3,
  title.line = 0,
  bounding.box = T,
  xlab = NULL,
  ylab = NULL,
  zlab = NULL,
  continuous.colors = NULL,
  continuous.color.limits = NULL,
  discrete.colors = colors.day
)
```

```{r plotDM_colTipsClusters}
tips_cells <- cellsInCluster(object, "day", "Day_11")
renamed_tips <- gsub("11_", "", object@group.ids[rownames(object@group.ids) %in% tips_cells, 'singleTP_res1'])
object@group.ids[rownames(object@group.ids) %in% tips_cells, "tip.clusters"] <- renamed_tips

object@group.ids$col_clusters <- ifelse(is.na(object@group.ids$tip.clusters), "not a tip cell", object@group.ids$tip.clusters)
colors.tips <- c(rainbow(length(sort(unique(object@group.ids$col_clusters)))-1), 'grey')
names(colors.tips) <- sort(unique(object@group.ids$col_clusters))
plotDimArray(object = object,
             reduction.use = "dm",
             dims.to.plot = 1:18,
             label="col_clusters",
             plot.title="",
             outer.title=paste0("tip.clusters - Diffusion Map Sigma ", sigma, "\n"),
             legend=F,
             alpha=0.3,
             discrete.colors=colors.tips
)

p1 <- plotDim(object, 'col_clusters', discrete.colors = colors.tips)
grid.newpage()
grid.draw(g_legend(p1))
```


```{r eval=FALSE, echo=FALSE}
plotDim3D(
  object,
  'tip.clusters',
  label.type = "search",
  reduction.use = c("dm"),
  view = NULL,
  dim.1 = 1,
  dim.2 = 2,
  dim.3 = 9,
  w.1 = NULL,
  w.2 = NULL,
  w.3 = NULL,
  cells = NULL,
  # cells = cellsInCluster(object, clustering = "tip.clusters", cluster = c("6","8","9","18")),
  alpha = 0.4,
  size = 4,
  title = NULL,
  title.cex = 3,
  title.line = 0,
  bounding.box = T,
  xlab = NULL,
  ylab = NULL,
  zlab = NULL,
  continuous.colors = NULL,
  continuous.color.limits = NULL,
  discrete.colors = colors.use
  # discrete.colors = colors.use[names(colors.use) %in% c("6","8","9","18")]
)
```

# Calculate pseudotime (PT)

```{r floodPT}
set.seed(seed)
flood.result <- floodPseudotime(object,
                                root.cells=root.cells,
                                n=floodNsim,
                                minimum.cells.flooded=2,
                                verbose=F
)
saveRDS(flood.result, file = paste0(PATH_RDS, "flood_Nsim.", floodNsim, ".rds"))
```


```{r floodPTprocess}
set.seed(seed)
object <- floodPseudotimeProcess(object,
                                 flood.result,
                                 floods.name="pseudotime",
                                 max.frac.NA=0.4,
                                 pseudotime.fun=mean,
                                 stability.div=20
)

pseudotimePlotStabilityOverall(object)
plotDimArray(object = object,
             reduction.use = "dm",
             dims.to.plot = 1:18,
             label = "pseudotime",
             plot.title = "",
             outer.title = "Diffusion Map labeled by pseudotime",
             legend = F,
             alpha = 0.4
)

plotDim(object, reduction.use = "tsne", "pseudotime", plot.title = "Pseudotime") +
    xlab("UMAP_1") +
    ylab("UMAP_2")
plotDists(object, "pseudotime", "day", plot.title = "Pseudotime by day")
```

```{r savePT}
saveRDS(object,
        file = paste0(PATH_RDS, "object_withDMandPT.rds")
)
```

```{r setupBis, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  echo = TRUE
)
```

# Define tips

```{r tipsPlot}
plotDim(
    object,
    "col_clusters",
    discrete.colors = colors.tips,
    reduction.use = "tsne",
    plot.title = "Leiden clusters -\nfrom Seurat analysis",
    legend = FALSE) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank()
    ) +
    xlab("UMAP_1") +
    ylab("UMAP_2")

p1 <- plotDim(object, 'col_clusters', discrete.colors = colors.tips)
grid.newpage()
grid.draw(g_legend(p1))
```

# Biased random walks

```{r performPTlogistic}
set.seed(seed)
object.ptlogistic <- pseudotimeDetermineLogistic(object,
                                                 "pseudotime",
                                                 optimal.cells.forward=cellsForward,
                                                 max.cells.back=cellsBack,
                                                 do.plot = T
)

# Bias the transition matrix according to pseudotime
set.seed(seed)
object.biased.tm <- as.matrix(pseudotimeWeightTransitionMatrix(object,
                                                               "pseudotime",
                                                               logistic.params=object.ptlogistic
))

# Simulate the biased random walks from each tip
set.seed(seed)
object.walks <- simulateRandomWalksFromTips(object,
                                            tip.group.id="tip.clusters",
                                            root.cells=root.cells,
                                            transition.matrix = object.biased.tm,
                                            max.steps = RWmaxSteps,
                                            n.per.tip = ifelse(RWNsim == 2, 2 * length(tips_cells), RWNsim),
                                            verbose = F
)

# Process the biased random walks into visitation frequencies
set.seed(seed)
object <- processRandomWalksFromTips(object, object.walks, verbose = F)

saveRDS(object, file = paste0(PATH_RDS, "object_withRW_cellsBandF.", cellsForward, "-", cellsBack, "_RWmaxSteps.", RWmaxSteps, "_RWNsim.", ifelse(RWNsim == 2, 2 * length(tips_cells), RWNsim), ".rds"))
```

# Build tree

```{r buildTree}
# Load the cells used for each tip into the URD object
object.tree <- loadTipCells(object, "tip.clusters")
# Build the tree
set.seed(seed)
object.tree <- buildTree(object.tree,
                         tips.use = levels(as.factor(object.tree@group.ids$tip.clusters)),
                         pseudotime = "pseudotime",
                         divergence.method = treeMeth,
                         cells.per.pseudotime.bin = cellsPerPTbin,
                         bins.per.pseudotime.window = binsPerPTwindow,
                         save.all.breakpoint.info = T,
                         p.thresh = treePtreshold,
                         min.cells.per.segment = 0
)
```

# Vizualisation {.tabset .tabset-pills .tabset-fade}

## General information

```{r plot-generalInfo}
plotTree(object.tree, title="Labelled segments", tree.size = 0.3, label.segments = T)

plotTree(object.tree, "day", title="Developmental Stage", tree.size = 0.3) +
    theme(plot.title = element_text(hjust = 0.5))
```

## Clusters from the single time point clustering

```{r plot-clusters, results='asis', out.width='20%', fig.align='default'}
# loop over the clusters of singleTP_res1
for (cluster in levels(factor(object.tree@group.ids$singleTP_res1, levels = c(
    "4_1", "4_2", "4_3", "4_4", "4_5", "4_6", "4_7", "4_8", "4_9", "4_10", "4_11", "4_12", "4_13", "4_14", "4_15",
    "5_1", "5_2", "5_3", "5_4", "5_5", "5_6", "5_7", "5_8", "5_9", "5_10", "5_11", "5_12", "5_13", "5_14",
    "6_1", "6_2", "6_3", "6_4", "6_5", "6_6", "6_7", "6_8", "6_9", "6_10", "6_11", "6_12", "6_13", "6_14", "6_15", "6_16",
    "11_1", "11_2", "11_3", "11_4", "11_5", "11_6", "11_7", "11_8", "11_9", "11_10", "11_11", "11_12", "11_13", "11_14", "11_15", "11_16", "11_17", "11_18", "11_19", "11_20"
)))){
    
    if (!grepl("^4_", cluster)){
        # create new group.id with the cluster only
        cellsCluster <- cellsInCluster(object.tree, "singleTP_res1", cluster)
        object.tree@group.ids[rownames(object.tree@group.ids) %in% cellsCluster, "clusterOI"] <- object.tree@group.ids[rownames(object.tree@group.ids) %in% cellsCluster, "singleTP_res1"]
        
        # plot the cluster of interest
        print(plotTree(object.tree, "clusterOI", title = cluster, tree.size = 0.3, cells.highlight = cellsCluster) +
                  theme(plot.title = element_text(hjust = 0.5)))
        
        # remove the metadata
        object.tree@group.ids$clusterOI <- NULL
    }
}

```

## Gene of interest

```{r plot-goi, results='asis', out.width='20%', fig.align='default'}
# Load list of genes of interest
goi <- read.table("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/inputData/other/genes_of_interest3-scRNAseq.tsv", sep = ",")
goiList <- goi$V1

for (gene in goiList){
    print(plotTree(object.tree, gene, title = gene, tree.size = 0.3, cells.highlight = whichCells(object.tree, gene, c(0.1, 1000)), cells.highlight.size = 0.5) +
                  theme(plot.title = element_text(hjust = 0.5)))
}
```

# Save object

```{r}
saveRDS(object.tree, file = paste0(PATH_RDS, "object_tree_METH.", treeMeth, "_cellsPerPTbin.", cellsPerPTbin, "binsPerPTwindow.", binsPerPTwindow, ".rds"))
```

```{r, eval=FALSE, echo=FALSE}
object.tree <- readRDS(file = paste0(PATH_RDS, "object_tree_METH.", treeMeth, "_cellsPerPTbin.", cellsPerPTbin, "binsPerPTwindow.", binsPerPTwindow, ".rds"))
object <- readRDS(paste0(PATH_RDS, "object_withRW_cellsBandF.", cellsForward, "-", cellsBack, "_RWmaxSteps.", RWmaxSteps, "_RWNsim.", ifelse(RWNsim == 2, 2 * length(tips_cells), RWNsim), ".rds"))
shortTree <- loadTipCells(object, "tip.clusters")
tips.to.build <- c("16","18","8","9","6","4", "14")

set.seed(seed)
shortTree <- buildTree(shortTree,
                         tips.use = tips.to.build,
                         pseudotime = "pseudotime",
                         divergence.method = treeMeth,
                         cells.per.pseudotime.bin = cellsPerPTbin,
                         bins.per.pseudotime.window = binsPerPTwindow,
                         save.all.breakpoint.info = T,
                         p.thresh = treePtreshold,
                         min.cells.per.segment = 0
)
```

```{r plotDual, eval=FALSE, echo=FALSE}
plotTree(shortTree, "day")
plotTreeDual(object.tree, "Pax3", "Tbx1")
```

