---
author: "Céline Chevalier, Anaïs Baudot, Fabienne Lescroart, Laurent Argiro, Stéphane Zaffran"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
title: |
  | lab_day_11 : 
  | specific cardiomyocytes markers
---


<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    });
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    });
  });
</script>

```{r setup, include=FALSE, eval=TRUE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # text output
  echo = TRUE,
  results = 'hold',
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,
  
  # code decoration
  #tidy = TRUE,
  tidy.opts = list(width.cutoff = 90),
  comment = '',
  attr.output='.numberLines',
  
  # plots
  #fig.path = "figure/",      # is set later, in chunk setup2
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c('png', 'pdf'),
  fig.width = 8,
  fig.height = 8,
  #fig.asp=1.3               # fig.height = fig.width * fig.asp
  #out.extra = 'style="border:5px solid orange"'     # orange box arround plots
  fig.align = "center"       # should be tuned to default in multiple plots chunk
)
```

```{r setup2, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # figures
  fig.path = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/ciml5/lab_day_11/figures/"
)
```

```{r setupAnalysis, eval=TRUE, include=FALSE}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(patchwork))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(gridExtra))

source("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/gastruloid_timeserie_scRNA-seq/scripts/utilities/extract_topn.R")
source("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/gastruloid_timeserie_scRNA-seq/scripts/utilities/add_dominantMetadata.R")
source("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/gastruloid_timeserie_scRNA-seq/scripts/utilities/00_generalDeps.R")

dataset <- "lab_day_11"

colors.table <- read.table("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/inputData/InputTables/ClusterColors.tsv", sep="\t", header=T, comment.char="", as.is=T)
colors.celltype <- setNames(colors.table$blind_friendly[!is.na(colors.table$transferred_identity)], colors.table$transferred_identity[!is.na(colors.table$transferred_identity)])
colors.stage <- setNames(c("#bb573b", "#ac9d3e", "#5aaf6a", "#7564b9", "#b94c84", "#ba4b7d", "#90b03d", "#60a756", "#45c097", "#5e8bd5", "#6d71d8", "#573585", "#bd80d5", "#b853a2", "#ba4b7d"), c("Day_04", "Day_05", "Day_06", "Day_07", "Day_10", "Day_11", "E6.5", "E6.75", "E7.0", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5"))
```


# Context

```{r loadData}
do.scale <- FALSE
pca.npcs <- 50
top.pcs <- 30
general.seed <- 17
top.markers <- 10

SO <- readRDS("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/ciml5/rdsObjects/03_WF_onSinglets_lab_day_11.rds")
celltypes <- read.table(
    file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/ciml5/lab_day_11/cellIdentitiesPredictions_final_lab_day_11.csv",
    sep = ",",
    header = TRUE)
umap2d <- read.table(
    file = "/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/ciml5/lab_day_11/final_umap2Dcoordinates_lab_day_11.csv",
    sep = ",",
    header = TRUE)
SO[["umap2d"]] <- CreateDimReducObject(embeddings = as.matrix(umap2d), key = "UMAP2d_")
SO <- AddMetaData(SO, metadata = celltypes$predicted.id, col.name = "celltype")
```


```{r overview, out.width='50%', fig.show='hold'}
clusterResName <- "RNA_snn_res.1"
Idents(SO) <- clusterResName
clustersColor <- setNames(hue_pal()(length(levels(SO@meta.data[,clusterResName]))), levels(SO@meta.data[,clusterResName]))
pover1 <- DimPlot(
    SO,
    reduction = "umap2d",
    pt.size = 0.8,
    label = FALSE,
    cols = clustersColor[levels(Idents(SO))]) +
    NoLegend() +
    ggtitle(paste0(dataset, " : clusters at resolution 1")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())
pover1 <- LabelClusters(pover1, id = "ident",  fontface = "bold", size = 5)


Idents(SO) <- SO@meta.data$celltype
pover2 <- DimPlot(
    SO,
    reduction = "umap2d",
    pt.size = 0.8,
    label = FALSE) +
    ggtitle(paste0(dataset, " : celltype identities")) +
    NoLegend() +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())
#pover2 <- LabelClusters(pover2, id = "ident",  fontface = "bold", size = 5)

p1 <- DimPlot(SO, reduction = "umap2d")
p2 <- cowplot::get_legend(p1)

pover1
pover2
grid.newpage()
grid.draw(p2)
```

# Subset to cluster 16 {.tabset  .tabset-pills .tabset-fade}

```{r dimplot-16, out.width='50%'}
cellsData <- data.frame(SO@reductions[["umap2d"]]@cell.embeddings, SO@meta.data$RNA_snn_res.1)
colnames(cellsData) <- c(colnames(SO@reductions[["umap2d"]]@cell.embeddings), "cluster")

fullPlot <- ggplot(cellsData[c('UMAP2d_1', 'UMAP2d_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP2d_1,
            y = UMAP2d_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$cluster %in% c(16),], # Now provide data including column for facetting
                aes(color = cluster),
                size  = 1.2) +
    scale_fill_manual(values = clustersColor[c(16)], aesthetics = "color") +
    theme_void() +
    ggtitle(paste0(dataset, " : cluster 16")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank(),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size=8, alpha = 1))) +
    NoLegend()
fullPlot
```

```{r}
cl16 <- subset(SO, cells = rownames(SO@meta.data[which(SO@meta.data$RNA_snn_res.1 == 16),]))
cellsData <- data.frame(cl16@reductions[["umap2d"]]@cell.embeddings, cl16@meta.data$RNA_snn_res.1)
colnames(cellsData) <- c(colnames(cl16@reductions[["umap2d"]]@cell.embeddings), "cluster")

cluster16plot <- ggplot(cellsData[c('UMAP2d_1', 'UMAP2d_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP2d_1,
            y = UMAP2d_2)) +
    geom_point( data = cellsData[cellsData$cluster %in% c(16),], # Now provide data including column for facetting
                aes(color = cluster),
                size  = 1.2) +
    scale_fill_manual(values = clustersColor[c(16)], aesthetics = "color") +
    theme_void() +
    ggtitle(paste0(dataset, " : zoom on the cluster 16")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank(),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size=8, alpha = 1))) +
    NoLegend()

cluster16plot
```

## Expressed factors : Tbx1 and Pax3 {.tabset .tabset-pills .tabset-fade}

### Table

```{r}
# create metadata for expression of Tbx1
cl16@meta.data["Tbx1+"] <- cl16@assays$RNA@data["Tbx1",] != 0
# create metadata for expression of Pax3
cl16@meta.data["Pax3+"] <- cl16@assays$RNA@data["Pax3",] != 0
# Create a table
coExprMtx <- as.data.frame.matrix(table(cl16@meta.data$`Tbx1+`, cl16@meta.data$`Pax3+`))
coExprMtx %>%
    knitr::kable(caption = "Coexpression of the genes *Tbx1* and *Pax3* ?", 
                 align = "c") %>% 
    kable_paper(full_width = F) %>%
    add_header_above(c(" " = 1, "Pax3+" = 2)) %>%
    pack_rows("Tbx1+", 1, 2)
```

<!-- ### DimPlot -->

```{r}
# Create a plot
cellsData <- data.frame(cl16@reductions[["umap2d"]]@cell.embeddings, cl16@meta.data$`Tbx1+`, cl16@meta.data$`Pax3+`)
colnames(cellsData) <- c(colnames(cl16@reductions[["umap2d"]]@cell.embeddings), "Tbx1+", "Pax3+")

geneCol <- hue_pal()(2)
cluster16plot <- ggplot(cellsData[c('UMAP2d_1', 'UMAP2d_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP2d_1,
            y = UMAP2d_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$`Tbx1+` == TRUE,], # Now provide data including column for facetting
                aes(color = "Tbx1+"),
                size  = 1.2) +
    geom_point( data = cellsData[cellsData$`Pax3+` == TRUE,], # Now provide data including column for facetting
                aes(color = "Pax3+"),
                size  = 1.2) +
    scale_fill_manual(values = geneCol, aesthetics = "color") +
    theme_void() +
    ggtitle(paste0(dataset, " : cluster 16")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank(),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size=8, alpha = 1)))
cluster16plot

# leg2 <- cowplot::get_legend(cluster16plot)
# cluster16plot + NoLegend()
# grid.newpage()
# grid.draw(leg2)
```


# Heapmat with hierarchical clustering

```{r}
library(pheatmap)
dim(cl16@assays$RNA@data)

"Tbx1" %in% rownames(cl16@assays$RNA@data[rowMeans(cl16@assays$RNA@data) > 0.05, ])
dim(cl16@assays$RNA@data[rowMeans(cl16@assays$RNA@data) > 0.05, ])
cl16expr <- cl16@assays$RNA@data[rowMeans(cl16@assays$RNA@data) > 0.05, ]

cl16marker <- ifelse(cl16@meta.data$`Tbx1+` == TRUE, "Tbx1", ifelse(cl16@meta.data$`Pax3+` == TRUE, "Pax3", "none"))

heatmap <- pheatmap(cl16expr, show_rownames = FALSE, show_colnames = FALSE, cutree_rows = 2)
heat_cluster <- as.data.frame(cutree(heatmap$tree_row, k = 2))
colnames(heat_cluster) <- "treeCluster"

rownames(heat_cluster)[which(heat_cluster$treeCluster==2)]
```

Observation on the heatmap  
*(columns: cells, rows: features)*
The cells do not seem cluster. The pattern of expression of the cells looks
equivalent among all cells.

Observation of the clusters with highest expression
The gene list of the most expressed genes in the cluster 16 are mainly related
to the ribosomal acitvity (`Rp(ls)`).

# Supp : FeaturePlots of the genes of interest of Day_11

```{r featurePlot-18versus16, out.width='20%', fig.align='default'}
genes <- c("Myod1", "Myog", "Myl7", "Sox7", "Cald1", "Acta2", "Vsnl1", 
           "Nppa", "Itga6", "Myh6", "Myl2", "Myh7", "Myl3",
           "Tbx18", "Wt1", "Mab21l2", "Myf5", "Pax7", "Met", "Myh3",
           "Isl1", "Pitx2", "Msc", "Tbx1", "Tcf21",
           "Pax3", "Meox1", "Tcf15", "Itm2a")
table(genes %in% rownames(cl16))
invisible( lapply( genes, function(featureName) {
    print( FeaturePlot( cl16, features = featureName, reduction = ifelse( exists("useReduction"), useReduction, "umap2d"), order = TRUE, min.cutoff = "q1", max.cutoff = "q99") +
               theme_void() +
               NoLegend() +
               CenterTitle())
               # theme( axis.title.x = element_blank(),
               #        axis.title.y = element_blank(),
               #        legend.position = "none"))
}))
```
