---
author: "Céline Chevalier, Anaïs Baudot, Fabienne Lescroart, Laurent Argiro, Stéphane Zaffran"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
params:
    dataset:
        value: NA
        choices: [lab_day_04, lab_day_05, lab_day_06, lab_day_11, lab_4_days, lab_3_days]
    analysis.name:
        value: "ciml5"
    general.seed: 17
title: |
  | Protocole 3:
  | Data preparation for Python analyses
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    })
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    })
  })
</script>



```{r setup1, eval=TRUE, include=FALSE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # text output
  echo = TRUE,
  results = 'hold',
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,
  
  # code decoration
  #tidy = TRUE,
  tidy.opts = list(width.cutoff = 90),
  comment = '',
  attr.output='.numberLines',
  
  # plots
  #fig.path = "figure/",      # is set later, in chunk setup2
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c('png', 'pdf'),
  fig.width = 8,
  fig.height = 8,
  #fig.asp=1.3               # fig.height = fig.width * fig.asp
  #out.extra = 'style="border:5px solid orange"'     # orange box arround plots
  fig.align = "center"       # should be tuned to default in multiple plots chunk
)
```

```{r load-lib, include=FALSE}
library(dplyr)
library(Matrix)
library(Seurat)
library(ggplot2)
library(reticulate)
library(scmap)
library(RColorBrewer)
library(DoubletFinder)
library(stringr)
library(kableExtra)
library(formattable)
library(knitr)
library(SingleCellExperiment)
library(BiocGenerics)
library(scales)
library(formatR)
library(uwot)
library(DT)
library(patchwork)
library(ggrepel)
```

```{r get-params, include=FALSE}
bindingIsLocked("params", env = .GlobalEnv)
unlockBinding("params", env = .GlobalEnv)
attach(params)
set.seed(general.seed)
```

# Global settings {.tabset .tabset-pills .tabset-fade}

## Directories

Set either input and output directories.

```{r base-path}
basePath <- "../../../seuratAnalysis" # local server
#basePath <- "/shared/projects/mothard_in_silico_modeling/seurat_analysis" # IFB server

# Input directories
PATH_ALL_INPUT <- file.path(basePath, "inputData")
PATH_LABDATA <- file.path(PATH_ALL_INPUT, "lab_data")

PATH_BASE_ANALYSIS <- file.path(basePath, analysis.name)
PATH_RDS_OBJECTS <- file.path(PATH_BASE_ANALYSIS, "rdsObjects")

sep_data <- grep("^03_", list.files(PATH_RDS_OBJECTS, full.names = FALSE, recursive = FALSE), value = TRUE)
sep_days <- sort(str_extract(sep_data, "lab_day_[0-9]{2}"))
```

```{r starter, eval=TRUE, include=FALSE}
knitr::opts_chunk$set(
    # code evaluation
    eval = TRUE,
    echo = TRUE
)
```

# Load the filtered data

Take data previously generated during single dataset pipeline execution.

```{r multiple-load}
if (dataset == "lab_3_days") {
    
    sep_data.list <- lapply(sep_days[1:3], function(day){
        
        toLoad_SO <- grep(day, sep_data, value = TRUE)
        SO <- readRDS(file = file.path(PATH_RDS_OBJECTS, toLoad_SO))
    })
    names(sep_data.list) <- sep_days[1:3]
    
    
} else if (dataset == "lab_4_days") {
    
    sep_data.list <- lapply(sep_days, function(day){
        
        toLoad_SO <- grep(day, sep_data, value = TRUE)
        SO <- readRDS(file = file.path(PATH_RDS_OBJECTS, toLoad_SO))
    })
    names(sep_data.list) <- sep_days
} else if (dataset %in% sep_days) {
    
    sep_data.list <- list()
    sep_days <- grep(dataset, sep_days, value = TRUE)
    toLoad_SO <- grep(dataset, sep_data, value = TRUE)
    sep_data.list[[dataset]] <- readRDS(file = file.path(PATH_RDS_OBJECTS, toLoad_SO))
}

sep_data.list
```

# List features and cellnames of filtered in data

```{r}
invisible(sapply(sep_days, function(datasetX){
    cells <- colnames(sep_data.list[[datasetX]])
    features <- rownames(sep_data.list[[datasetX]])
    write(
        cells,
        file = file.path(PATH_BASE_ANALYSIS, datasetX, paste0("filteredIn_cells_", datasetX, ".txt")))
    write(
        features,
        file = file.path(PATH_BASE_ANALYSIS, datasetX, paste0("filteredIn_features_", datasetX, ".txt")))
}))
rm(sep_data.list)
gc()
```

# Load raw data

```{r}
rawData.list <- sapply(sep_days, function(datasetX){
    # Load filtered cells and features
    cellsIn <- readLines(file.path(PATH_BASE_ANALYSIS, datasetX, paste0("filteredIn_cells_", datasetX, ".txt")))
    featuresIn <- readLines(file.path(PATH_BASE_ANALYSIS, datasetX, paste0("filteredIn_features_", datasetX, ".txt")))
    
    # Load raw data
    raw.data <- Read10X(data.dir = file.path(PATH_LABDATA, datasetX), gene = 2)
    SO <- CreateSeuratObject(counts=raw.data, project=dataset)
    
    # rename cells
    cell_ids <- colnames(SO)
    UMIs <- str_extract(cell_ids, '[A-Z]*')
    cellnames <- paste(datasetX, 1, UMIs, sep = '_')
    SO <- RenameCells(SO, new.names = cellnames)
    
    # subset the raw data
    SO_subsetCells <- SO[featuresIn, cellsIn]
})
```

# Merge the data and normalize

```{r}
if (dataset == "lab_3_days" | dataset == "lab_4_days") {
    
    SO <- merge(x = rawData.list[[1]], y = rawData.list[2:length(rawData.list)], merge.data = T)

} else {
    
    SO <- rawData.list[[1]]
    
}

rm(sep_data, rawData.list)
invisible(gc())

SO <- NormalizeData(SO, normalization.method = "LogNormalize", verbose=FALSE)
```

# Save the object

```{r}
saveRDS(SO, file = file.path(PATH_RDS_OBJECTS, paste0("99_rawData_filtered_", dataset, ".rds")))
```


<!-- nohup Rscript -e "rmarkdown::render('cellrank_dataPreparation.Rmd', params = list( dataset = 'lab_4_days' ), output_file = './cellrank_dataPreparation_ciml5_lab_4_days.html')" & -->
