---
author: "Céline Chevalier, Anaïs Baudot"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
params:
    dataset:
        value: NA
        choices: [lab_day_04, lab_day_05, lab_day_06, lab_day_11]
title: |
  | Protocole 1.2:
  | `r params$dataset` integration to the atlas exploration
---


<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    });
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    });
  });
</script>

```{r setup, include=FALSE, eval=TRUE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # text output
  echo = TRUE,
  results = 'hold',
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,
  
  # code decoration
  tidy.opts = list(width.cutoff = 90),
  comment = '',
  attr.output='.numberLines',
  
  # plots
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c('png', 'pdf'),
  fig.width = 12,
  fig.height = 12,
  fig.align = "center"       # should be tuned to default in multiple plots chunk
)
```

```{r setupAnalysis, eval=TRUE, include=FALSE}
library(Seurat)
library(dplyr)
library(cowplot)
library(ggplot2)
library(patchwork)

bindingIsLocked("params", env = .GlobalEnv)
unlockBinding("params", env = .GlobalEnv)
attach(params)

colors.table <- read.table("../../../seuratAnalysis/inputData/InputTables/ClusterColors.tsv", sep="\t", header=T, comment.char="", as.is=T)
colors.celltype <- setNames(colors.table$blind_friendly[!is.na(colors.table$transferred_identity)], colors.table$transferred_identity[!is.na(colors.table$transferred_identity)])
colors.stage <- setNames(c("#bb573b", "#ac9d3e", "#5aaf6a", "#7564b9", "#b94c84", "#ba4b7d", "#90b03d", "#60a756", "#45c097", "#5e8bd5", "#6d71d8", "#573585", "#bd80d5", "#b853a2", "#ba4b7d"), c("Day_04", "Day_05", "Day_06", "Day_07", "Day_10", "Day_11", "E6.5", "E6.75", "E7.0", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5"))
```

```{r setup2, eval=TRUE, include=FALSE}
bindingIsLocked("params", env = .GlobalEnv)
unlockBinding("params", env = .GlobalEnv)
fig.folder <- file.path("../../../seuratAnalysis/ciml5", dataset, "figures/")
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # figures
  fig.path = fig.folder
)
```

# Load data

```{r loadAtlas}
do.scale <- FALSE
pca.npcs <- 50
top.pcs <- 30
general.seed <- 17

SO <- readRDS(paste0("../../../seuratAnalysis/ciml5/rdsObjects/03_WF_onSinglets_", dataset, ".rds"))
atlas.subset <- readRDS("../../../seuratAnalysis/inputData/atlas/atlas_preprocessed.rds")
celltypesPred <- read.table(
    paste0("../../../seuratAnalysis/ciml5/", dataset, "/cellIdentitiesPredictions_final_", dataset, ".csv"),
    sep = ",", header = TRUE)
```

# Perform integration

```{r results='asis', out.width="50%"}
features <- SelectIntegrationFeatures(object.list = list(SO, atlas.subset))
anchors <- FindIntegrationAnchors(object.list = list(SO, atlas.subset), anchor.features = features)
SO.align <- IntegrateData(anchorset = anchors)

SO.align@project.name <- paste0(dataset, "_with_atlas")
DefaultAssay(SO.align) <- "integrated"

SO.align <- ScaleData(SO.align, assay = DefaultAssay(SO.align), do.scale=do.scale, verbose = FALSE)
SO.align <- RunPCA(SO.align, npcs = pca.npcs, verbose = FALSE)
SO.align <- RunUMAP(SO.align, dims = 1:top.pcs, seed.use = general.seed, verbose = FALSE)
SO.align@meta.data[SO.align@meta.data$model == "Gastruloids", "celltype"] <- celltypesPred$predicted.id
```


# Segregation by model

```{r model-segregation}
Idents(SO.align) <- "model"
cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model")
pmodels <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP_1,
            y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData, # Now provide data including column for facetting
                aes(color = model),
                alpha = .5,
                size  = 1.2) +
    facet_wrap(facets = vars(factor(model, levels = c("Embryos", "Gastruloids"))),
               ncol = 2) +
    theme_void() +
    NoLegend() +
    ggtitle(paste0(dataset, " : atlas cells (as Embryos)\n and gastruloids after integration")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

pmodels
```


# Overview of the atlas and data integration

```{r overview}
Idents(SO.align) <- "model"
cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model, SO.align@meta.data$day)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model", "day")

pover1 <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$model == "Embryos",], # Now provide data including column for facetting
                aes(color = day),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.stage[cellsData[cellsData$model == "Embryos", "day"]],
                      aesthetics = "color",
                      breaks = sort(levels(as.factor(cellsData[cellsData$model == "Embryos", "day"]))),
                      name = "Stages") +
    theme_void() +
    ggtitle("Atlas stages") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank(),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size=8, alpha = 1)))

cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model, SO.align@meta.data$celltype)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model", "celltype")

pover2 <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$model == "Embryos",], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.celltype[SO.align@meta.data$celltype], aesthetics = "color") +
    theme_void() +
    NoLegend() +
    ggtitle("Atlas celltypes overview") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

pover3 <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$model == "Gastruloids",], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.celltype[SO.align@meta.data$celltype], aesthetics = "color") +
    theme_void() +
    NoLegend() +
    ggtitle(paste0(dataset, " predicted celltypes from the atlas transfer")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

layout <- "
AAABB
AAACC
"

pover1 + pover2 + pover3 +
    plot_layout(design = layout) +
    plot_annotation(title = paste0(
        "Visualization of the timestamps and the celltypes\nof the integrated ", dataset, " data with the atlas"),
        theme = theme(plot.title = element_text(hjust = 0.5))
        )
```

# Visualisations {.tabset  .tabset-pills .tabset-fade}

## Atlas stages

### Overview

```{r stagesAtlas-all}
pover1
```

### Facets

```{r stagesAtlas-facets}
cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model, SO.align@meta.data$day)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model", "day")
stageF <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP_1,
            y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$model == "Embryos",], # Now provide data including column for facetting
                aes(color = day),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.stage[cellsData[cellsData$model == "Embryos", "day"]],
                      aesthetics = "color",
                      breaks = sort(levels(as.factor(cellsData[cellsData$model == "Embryos", "day"]))),
                      name = "Stages") +
    facet_wrap(facets = vars(day),
               ncol = 3) +
    theme_void() +
    NoLegend() +
    ggtitle("Atlas stages") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

stageF
```

## Atlas celltypes

### Overview

```{r celltypesAtlas-all}
pover2
```

### Facets

Here only the common celltypes with the data are represented in the facets.

```{r celltypesAtlas-facets, fig.height = 16}
cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model, SO.align@meta.data$celltype)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model", "celltype")

celltypeAtlasF <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP_1,
            y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$celltype %in% cellsData[cellsData$model == "Gastruloids", "celltype"] & cellsData$model == "Embryos",], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.celltype[SO.align@meta.data$celltype], aesthetics = "color") +
    facet_wrap(facets = vars(celltype),
               ncol = 3) +
    theme_void() +
    NoLegend() +
    ggtitle("Atlas celltypes overview") +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

celltypeAtlasF
```


## Dataset celltypes

### Overview

```{r celltypesData-all}
celltypeDataAll <- ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$model == "Gastruloids",], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 1.2) +
    scale_fill_manual(values = colors.celltype[SO.align@meta.data$celltype], aesthetics = "color") +
    theme_void() +
    NoLegend() +
    ggtitle(paste0(dataset, " predicted celltypes from the atlas transfer")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())

celltypeDataAll
```

### Facets {.tabset .tabset-pills .tabset-fade}

Here,  I display all the celltypes present in the dataset, either on atlas and dataset cells.

```{r celltypesData-facet, results='asis'}
invisible(sapply(levels(as.factor(SO@meta.data$celltype)), function(act)
{
    cat("#### ", act, "\n")
    
    celltypeDataF <- ggplot(
        cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
        aes( x = UMAP_1,
             y = UMAP_2)) +
        geom_point( alpha = .4,
                    size  = 1,
                    color = "grey") +
        geom_point( data = cellsData[cellsData$celltype == act, ], # Now provide data including column for facetting
                    aes(color = celltype),
                    alpha = .5,
                    size  = 1.2) +
        scale_fill_manual(values = colors.celltype[SO.align@meta.data$celltype], aesthetics = "color") +
        facet_wrap(facets = vars(model),
                   ncol = 2) +
        theme_void() +
        NoLegend() +
        ggtitle(act) +
        theme(plot.title = element_text(hjust = 0.5),
              axis.text = element_blank(),
              line = element_blank())
    
    print(celltypeDataF)
    
    cat(" \n \n") # Required for '.tabset'
}))
```


## Celltypes profile

```{r celltypesProfile}
cellsData <- data.frame(SO.align@reductions[["umap"]]@cell.embeddings, SO.align@meta.data$model, SO.align@meta.data$celltype, SO.align@meta.data$day)
colnames(cellsData) <- c(colnames(SO.align@reductions[["umap"]]@cell.embeddings), "model", "celltype", "day")

transition <- table(cellsData$celltype,
                    cellsData$day,
                    cellsData$model)
dt <- as.data.frame(transition)
dt[dt == 0] <- NA
names(dt) <- c("celltype", "stage", "model", "freq")

p3 <- ggplot(dt[(dt$model == "Embryos" & dt$stage != unique(SO@meta.data$day)[1]) ,], aes(stage, celltype, fill=freq)) +
    geom_tile() +
    geom_text(aes(label = freq), color="black", size = 2.5, fontface="bold") +
    coord_fixed() +
    scale_fill_gradient(low = "white", high = "pink", na.value="white") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    guides(fill = guide_colourbar(barwidth = 0.5,
                                  barheight = 10,
                                  title = "Nb of cells")) +
    ggtitle("Number of cells by stage for each celltype") +
    CenterTitle() +
    labs(x = "Stages", y = paste0("Celltypes"))


p4 <- ggplot(dt[dt$model == "Gastruloids" & dt$stage == unique(SO@meta.data$day)[1],], aes(stage, celltype, fill=freq)) +
    geom_tile() +
    geom_text(aes(label = freq), color="black", size = 2.5, fontface="bold") +
    coord_fixed() +
    scale_fill_gradient(low = "white", high = "pink", na.value="white") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    guides(fill = guide_colourbar(barwidth = 0.5,
                                  barheight = 10,
                                  title = "Nb of cells")) +
    ggtitle("Number of cells by stage for each celltype") +
    CenterTitle() +
    labs(x = "Stages", y = paste0("Celltypes"))

p3 + p4
```

<!-- To render  -->

<!-- for i in lab_day_04 lab_day_05 lab_day_06 lab_day_11; do nohup Rscript -e "rmarkdown::render('singleIntegrationExploration.Rmd', params = list( dataset = '$i'), output_file = './singleIntegrationExploration_ciml5_$i.html')" ; done & -->






