---
title: "Atlas exploration"
author: "CC"
date: "2022-11-30"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
---


<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src="https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type="text/javascript">
  $(document).ready(function() {
    $('body').prepend("<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use 'img.zoom' and out.extra='class=\"zoom\"' in chunk to specify manually which chunk images can be zoomed
    $('img:not(.zoomImg)').filter(':not(a *)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src'));
      $('.zoomDiv').show();
    });
    // onClick function for hiding div
    $('img.zoomImg').click(function() {
      $('.zoomDiv').hide();
    });
  });
</script>

```{r setup, include=FALSE, eval=TRUE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # text output
  echo = TRUE,
  results = 'hold',
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,
  
  # code decoration
  #tidy = TRUE,
  tidy.opts = list(width.cutoff = 90),
  comment = '',
  attr.output='.numberLines',
  
  # plots
  #fig.path = "figure/",      # is set later, in chunk setup2
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c('png', 'pdf'),
  fig.width = 12,
  fig.height = 12,
  #fig.asp=1.3               # fig.height = fig.width * fig.asp
  #out.extra = 'style="border:5px solid orange"'     # orange box arround plots
  fig.align = "center"       # should be tuned to default in multiple plots chunk
)
```

```{r setupAnalysis}
library(harmony)
library(Seurat)
library(dplyr)
library(cowplot)
library(ggplot2)

colors.table <- read.table("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/inputData/InputTables/ClusterColors.tsv", sep="\t", header=T, comment.char="", as.is=T)
colors.celltype <- setNames(colors.table$blind_friendly[!is.na(colors.table$transferred_identity)], colors.table$transferred_identity[!is.na(colors.table$transferred_identity)])
colors.stage <- setNames(c("#bb573b", "#ac9d3e", "#5aaf6a", "#7564b9", "#b94c84", "#ba4b7d", "#90b03d", "#60a756", "#45c097", "#5e8bd5", "#6d71d8", "#573585", "#bd80d5", "#b853a2", "#ba4b7d"), c("Day_04", "Day_05", "Day_06", "Day_07", "Day_10", "Day_11", "E6.5", "E6.75", "E7.0", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5"))
```


# Load Atlas

```{r}
do.scale <- FALSE
pca.npcs <- 50
top.pcs <- 30
general.seed <- 17

atlas_subset <- readRDS("/mnt/DATA_4TB/projects/gastruloids_sc_Lescroart/analysis/seuratAnalysis/inputData/atlas/atlas_preprocessed_UMAP.rds")
```

# Stages {.tabset  .tabset-pills .tabset-fade}

```{r stagesDimplot, results='asis', out.width="50%"}
Idents(atlas_subset) <- "day"
DimPlot(atlas_subset,
        reduction = "umap",
        pt.size = 1,
        repel = TRUE,
        label = TRUE,
        label.size = 6,
        cols = colors.stage[levels(Idents(atlas_subset))]) +
  ggtitle("Atlas : stages") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text = element_blank(),
        line = element_blank()) +
        NoLegend()

cellsData <- data.frame(atlas_subset@reductions[["umap"]]@cell.embeddings, atlas_subset@meta.data$day)
colnames(cellsData) <- c(colnames(atlas_subset@reductions[["umap"]]@cell.embeddings), "day")

invisible(sapply(levels(as.factor(atlas_subset@meta.data$day)), function(act)
{
    cat("## ", act, "\n")
        
    print(ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = .75,
                color = "grey") +
    geom_point( data = cellsData[cellsData$day == act,], # Now provide data including column for facetting
                aes(color = day),
                alpha = .5,
                size  = 2) +
    theme_void() +
    scale_fill_manual(values = colors.stage[atlas_subset@meta.data$day], aesthetics = "color") +
    NoLegend() +
    ggtitle(paste0(act)) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())
    )

    cat(" \n \n") # Required for '.tabset'
}))

```


# Celltypes {.tabset  .tabset-pills .tabset-fade}


```{r celltypesDimplot, results='asis', out.width="50%"}
Idents(atlas_subset) <- "celltype"
DimPlot(atlas_subset,
        reduction = "umap",
        pt.size = 1,
        repel = TRUE,
        label = TRUE,
        label.size = 6,
        cols = colors.celltype[levels(Idents(atlas_subset))]) +
  ggtitle("Atlas : cell identities") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text = element_blank(),
        line = element_blank()) +
        NoLegend()

cellsData <- data.frame(atlas_subset@reductions[["umap"]]@cell.embeddings, atlas_subset@meta.data$celltype)
colnames(cellsData) <- c(colnames(atlas_subset@reductions[["umap"]]@cell.embeddings), "celltype")

invisible(sapply(levels(as.factor(atlas_subset@meta.data$celltype)), function(act)
{
    cat("## ", act, "\n")
        
    print(ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = .75,
                color = "grey") +
    geom_point( data = cellsData[cellsData$celltype == act,], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 2) +
    theme_void() +
    scale_fill_manual(values = colors.celltype[atlas_subset@meta.data$celltype], aesthetics = "color") +
    NoLegend() +
    ggtitle(paste0(act)) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text = element_blank(),
          line = element_blank())
    )

    cat(" \n \n") # Required for '.tabset'
}))

```

# Cells repartition

```{r}
day <- NULL
transition <- table(atlas_subset@meta.data[WhichCells(atlas_subset, idents = day), 'day'], atlas_subset@meta.data[WhichCells(atlas_subset, idents = day), 'celltype'])
dt <- as.data.frame(transition, row.names = names(atlas_subset@meta.data[WhichCells(atlas_subset, idents = day), 'celltype']))
dt[dt == 0] <- NA
names(dt) <- c("stage", "celltype", "freq")

ggplot(dt, aes(stage, celltype, fill=freq)) +
    geom_tile() +
    geom_text(aes(label = freq), color="black", size = 2.5, fontface="bold") +
    #scale_color_manual(values = c(NA, "black"), na.value = NA) +
    coord_fixed() +
    scale_fill_gradient(low = "white", high = "blue", na.value="white") +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    guides(fill = guide_colourbar(barwidth = 0.5,
                                  barheight = 10,
                                  title = "Nb of cells")) +
    ggtitle("Number of cells by stage for each celltype") +
    CenterTitle() +
    labs(x = "Stages", y = paste0("Celltypes"))
```


